---
output: 
  github_document
---

# Hubbard Brook Forest Analytics 

The `HubbardBrookForestAnalytics` package (HBFA) facilitates easy access and consistent application of the revised Hubbard Brook Experimental Forest tree biomass equations. The associated Shiny app is available [here](https://kearutherford.shinyapps.io/HBFAShinyApp/). Further details on the revised biomass equations are provided in the accompanied manuscript:

Rutherford, K.H., J.J. Battles, and N.L. Cleavitt. 2026. Quantifying tree biomass: Insights from a long-term study at Hubbard Brook Experimental Forest. Canadian Journal of Forest Research.

> :bulb: **Tip:** you can navigate this README file using the table of contents found in the upper right-hand corner.

## Installation instructions

To install the `HubbardBrookForestAnalytics` package from GitHub:

```r
# install and load devtools
install.packages("devtools")
library(devtools)
```

```r
# install and load HubbardBrookForestAnalytics
devtools::install_github('kearutherford/HubbardBrookForestAnalytics')
library(HubbardBrookForestAnalytics)
```

## Citation instructions 

```{r}
citation("HubbardBrookForestAnalytics")
```

# Tree biomass estimates: `HBEFBiomass( )`

### Inputs

1. `data_type` Specifies whether the data of interest are internal to the package (meaning the clean, archived tree data from Hubbard Brook Experimental Forest for all watersheds, plots, and years) or external to the package (meaning data provided by the user). Must be set to "internal" or "external". The default is set to "internal".

2. `external_data` Only required if data_type is set to "external." A dataframe or tibble. Each row must be an observation of an individual tree. Must have at least these columns (column names are exact): 

    * **watershed:** Must be a character variable. Describes the watershed in which the data were collected. 
    
    * **year:** Must be a character variable. Describes the year in which the data were collected. 
    
    * **plot:** Must be a character variable. Identifies the plot in which the individual tree was measured.
    
    * **forest_type:** This column is OPTIONAL. If included, must be a character variable. Identifies the type of forest where the plot is located (for Hubbard Brook Experimental Forest, forest type would be either northern hardwood or spruce-fir).
        
    * **elev_m:** Must be a numeric variable. Elevation in meters above sea level for the plot. 
    
    * **hli:** Must be a numeric variable. Heat Load Index for the plot. Must be a value between -2.7 and 0.6.
    
    * **steep_deg:** Must be a numeric variable. Steepness (or slope) in degrees for the plot. Must be a value between 0 and 90.
            
    * **exp_factor:** Must be a numeric variable. The expansion factor specifies the number of trees per hectare that a given plot tree represents. 
    
    * **species:** Must be a character variable. Specifies the species of the individual tree. Must follow our four-letter species code naming conventions (see "Species code tables" in "Background information for tree biomass estimations" below).
    
    * **status:** Must be a character variable. Specifies whether the individual tree is alive ("Live") or dead ("Dead").
    
    * **vigor:** Must be a character variable. For live trees, the vigor should be 0, 1, 2, or 3 (see "Vigor code table" section in "Background information for tree biomass estimations" below). For standing dead trees, the vigor should be 4 or 5.
    
    * **dbh_cm:** Must be a numeric variable. Provides the diameter at breast height (DBH) of the individual tree in centimeters. 

3. `results` Specifies whether the results will be summarized by tree, by plot, or by plot as well as size class (size class has two categories: (1) tree, DBH >= 10 cm and (2) sapling, DBH < 10 cm). Must be set to either "by_tree", "by_plot", or "by_size". The default is set to "by_plot".

*Note: For plots without trees, species should be set to NONE. Watershed, year, plot, elev_m, and exp_factor columns should be filled in as appropriate. And all tree-level variables (status, vigor, dbh_cm) should be NA.*

### Outputs

A dataframe. The columns of the output dataframe depend on the results setting. 

* by_tree: 
    * All columns described above in external_data. 
    * **ht_cm:** Predicted height of the individual tree in centimeters.
    * **above_kg:** Aboveground tree biomass (including all wood, bark, branch, and foliage) in kilograms.
    * **leaf_kg:** Tree foliage biomass in kilograms.
    * Any additional (not required) columns in the input dataframe will remain in the output dataframe. 

* by_plot: 
    * **watershed:** As described above in external_data.
    * **year:** As described above in external_data.
    * **plot:** As described above in external_data.
    * **forest_type:** As described above in external_data.
    * **elev_m:** As described above in external_data.
    * **above_L_Mg_ha:** Aboveground live tree biomass (including all wood, bark, branch, and foliage) in megagrams per hectare.
    * **above_D_Mg_ha:** Aboveground dead tree biomass (including all wood, bark, branch) in megagrams per hectare.
    * **above_Mg_ha:** Aboveground (live + dead) tree biomass (including all wood, bark, branch, and foliage) in megagrams per hectare.
    * **leaf_Mg_ha:** Tree foliage biomass in megagrams per hectare.
  
* by_size: 
    * All columns described above in by_plot.
    * **sample_class:** Either tree (DBH >= 10 cm) or sapling (DBH < 10 cm).

### Demonstrations

```{r, include = FALSE}
library(HubbardBrookForestAnalytics)
```

**Results summarized by tree (using internal data):**
```{r}
# call the HBEFBiomass() function
internal_demo_1 <- HBEFBiomass(data_type = "internal",
                               results = "by_tree")

head(internal_demo_1)
```

*Notice in the output dataframe: The `aspect_deg`, `bbd`, `canopy`, `forest_type`, and `plot_area_ha` columns, which are additional (not required) columns in the internal dataframe, remain in the output dataframe. Any additional columns in the internal or external dataframe will remain in the output dataframe.*

<br>

**Results summarized by plot (using external data):**
```{r}
# investigate input external dataframe
external_demo_data
```

*Notice in the input dataframe: The plot without trees (W6, 1997, plot 2) is represented by species = "NONE". Watershed, year, plot, elev_m, and exp_factor are filled in as appropriate. All tree-level variables (i.e., status, vigor, dbh_cm) are NA.*

<br>

```{r}
# call the HBEFBiomass() function
internal_demo_2 <- HBEFBiomass(data_type = "external",
                               external_data = external_demo_data,
                               results = "by_plot")

internal_demo_2
```

*Notice in the output dataframe: The plot without trees (W6, 1997, plot 2) has 0 Mg/ha in all biomass columns.*

<br>

**Results summarized by plot as well as by size (using internal data):**
```{r}
# call the HBEFBiomass() function
internal_demo_3 <- HBEFBiomass(data_type = "internal",
                               results = "by_size")

head(internal_demo_3)
```

*Notice in the output dataframe: Since forest_type is included in the internal data, it is carried through into the summarized output data. The same idea applies to external data - if forest_type is included in the input dataframe, it will be carried through into the summarized output data.*

<br>

# Background information for tree biomass estimations

## Biomass workflow

**Step 1: Predict tree heights**

We fit new equations to predict height based on species, DBH, elevation, Heat Load Index, and steepness. These equations were fit using valley-wide height samples. 

<br>

**Step 2. Calculate aboveground live tree biomass**

We fit new equations to predict total aboveground tree biomass (including all wood, bark, branch, and foliage). We fit separate equations to predict tree foliage biomass on its own. See Rutherford et al. 2026 for details on the development of the biomass equations.

<br>

**Step 3. Discount aboveground tree biomass for dead trees**

Standing dead trees lose mass in two ways: (1) they degrade with pieces falling and transferring to other biomass pools (e.g., stem tops break and become coarse woody debris) and (2) the remaining structures decay as measured by their density (i.e., mass/volume). We accounted for the decay and degrade of standing dead trees by discounting the live biomass estimates based on vigor. For trees/saplings with a vigor code of 4, the discount rate is 0.7283 (aboveground live tree biomass x 0.7283). For trees/saplings with a vigor code of 5, the discount rate is 0.5683 (aboveground live tree biomass x 0.5683). 


## Species code tables 

All hardwood and softwood species currently included/recognized in the `HBEFBiomass()` function are listed in the tables below. 

**Hardwoods** 

|common name|scientific name|4-letter code|
|:---|:---|:---|
|Striped maple|Acer pensylvanicum|ACPE|
|Red maple|Acer rubrum|ACRU|
|Sugar maple|Acer saccharum|ACSA|
|Mountain maple|Acer spicatum|ACSP|
|Juneberry/serviceberry|Amelanchier spp.|AMSP|
|Yellow birch|Betula alleghaniensis|BEAL|
|Mountain birch|Betula cordifolia|BECO|
|Paper birch|Betula papyrifera|BEPA|
|Grey birch|Betula populifolia|BEPO|
|Alternateleaf dogwood|Cornus alternalternifolia|COAL|
|American beech|Fagus grandifolia|FAGR|
|White ash|Fraxinus americana|FRAM|
|Black ash|Fraxinus nigra|FRNI|
|Eastern hophornbeam|Ostrya virginiana|OSVI|
|Bigtooth aspen|Populus grandidentata|POGR|
|Quaking aspen|Populus tremuloides|POTR|
|Pin cherry|Prunus pensylvanica|PRPE|
|Black cherry|Prunus serotina|PRSE|
|Cherry species|Prunus spp.|PRSP|
|Chokecherry|Prunus virginiana|PRVI|
|Red oak|Quercus rubra|QURU|
|Willow species|Salix spp.|SASP|
|Red elderberry|Sambucus racemosa|SAPU|
|American mountain-ash|Sorbus americana|SOAM|
|Basswood|Tilia americana|TIAM|
|Unknown|NA|UNKN|

<br>

**Softwoods**

|common name|scientific name|4-letter code|
|:---|:---|:---|
|Balsam fir|Abies balsamea|ABBA|
|Red spruce|Picea rubens|PIRU|
|White pine|Pinus strobus|PIST|
|Eastern hemlock|Tsuga canadensis|TSCA|
|Unknown|NA|UNKN|

*Note: Four-letter species codes are generally the first two letters of the genus followed by the first two letters of the species.*


## Vigor code table 

|vigor class|description|
|:---|:---|
|0|alive (Siccama code for live)|
|1|alive, healthy|
|2|alive, sick|
|3|alive, dying|
|4|standing dead tree|
|5|standing dead tree with at least 1/3 of top missing (stub)|
|6|dead and down (not recognized in R package)|

*Note: There is some historical variance in vigor codes. These are Cleavitt-Battles era codes.*

