---
output: 
  github_document
---

# Hubbard Brook Forest Analytics 

The `HubbardBrookForestAnalytics` package (HBFA) ... 

> :bulb: **Tip:** you can navigate this README file using the table of contents found in the upper right-hand corner.

## Installation instructions

To install the `HubbardBrookForestAnalytics` package from GitHub:

```r
# install and load devtools
install.packages("devtools")
library(devtools)
```

```r
# install and load HubbardBrookForestAnalytics
devtools::install_github('kearutherford/HubbardBrookForestAnalytics')
library(HubbardBrookForestAnalytics)
```

## Citation instructions 

```{r}
citation("HubbardBrookForestAnalytics")
```

## Copyright notice

UC Berkeley's Standard Copyright and Disclaimer Notice: 

Copyright ©2024. The Regents of the University of California (Regents). All Rights Reserved. Permission to use, copy, modify, and distribute this software and its documentation for educational, research, and not-for-profit purposes, without fee and without a signed licensing agreement, is hereby granted, provided that the above copyright notice, this paragraph and the following two paragraphs appear in all copies, modifications, and distributions. 

IN NO EVENT SHALL REGENTS BE LIABLE TO ANY PARTY FOR DIRECT, INDIRECT, SPECIAL, INCIDENTAL, OR CONSEQUENTIAL DAMAGES, INCLUDING LOST PROFITS, ARISING OUT OF THE USE OF THIS SOFTWARE AND ITS DOCUMENTATION, EVEN IF REGENTS HAS BEEN ADVISED OF THE POSSIBILITY OF SUCH DAMAGE. 

REGENTS SPECIFICALLY DISCLAIMS ANY WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE. THE SOFTWARE AND ACCOMPANYING DOCUMENTATION, IF ANY, PROVIDED HEREUNDER IS PROVIDED "AS IS". REGENTS HAS NO OBLIGATION TO PROVIDE MAINTENANCE, SUPPORT, UPDATES, ENHANCEMENTS, OR MODIFICATIONS.

First published: July 1, 2024

# Tree biomass estimates

The `HBEFBiomass` function uses localized equations to estimate above-ground tree biomass. See "Background information for tree biomass estimations" below for further details. 

## :eight_spoked_asterisk: `HBEFBiomass( )`

### Inputs

1. `data_type` Specifies whether the data of interest are internal to the package (meaning the clean, archived tree data from Hubbard Brook Experimental Forest for all watersheds, plots, and years) or external to the package (meaning data provided by the user). Must be set to "internal" or "external". The default is set to "internal".

2. `external_data` Only required if data_type is set to "external." A dataframe or tibble. Each row must be an observation of an individual tree. Must have at least these columns (column names are exact): 

    * **watershed:** Must be a character variable. Describes the watershed in which the data were collected. 
    
    * **year:** Must be a character variable. Describes the year in which the data were collected. 
    
    * **plot:** Must be a character variable. Identifies the plot in which the individual tree was measured.
    
    * **forest_type:** This column is OPTIONAL. If included, must be a character variable. Identifies the type of forest where the plot is located (for Hubbard Brook Experimental Forest, forest type would be either northern hardwood or spruce-fir).
        
    * **elev_m:** Must be a numeric variable. Elevation in meters above sea level for the plot. 
            
    * **exp_factor:** Must be a numeric variable. The expansion factor specifies the number of trees per hectare that a given plot tree represents. 
    
    * **species:** Must be a character variable. Specifies the species of the individual tree. Must follow our four-letter species code naming conventions (see "Species code tables" in "Background information for tree biomass estimations" below).
    
    * **status:** Must be a character variable. Specifies whether the individual tree is alive ("Live") or dead ("Dead").
    
    * **vigor:** Must be a character variable. For live trees, the vigor should be 0, 1, 2, or 3 (see "Vigor code table" section in "Background information for tree biomass estimations" below). For standing dead trees, the vigor should be 4 or 5.
    
    * **dbh_cm:** Must be a numeric variable. Provides the diameter at breast height (DBH) of the individual tree in centimeters. 

3. `results` Specifies whether the results will be summarized by tree, by plot, or by plot as well as size class (size class has two categories: (1) tree, DBH >= 10 cm and (2) sapling, DBH < 10 cm). Must be set to either "by_tree", "by_plot", or "by_size". The default is set to "by_plot". 

### Outputs

A dataframe. The columns of the output dataframe depend on the results setting. 

* by_tree: 
    * All columns described above in external_data. 
    * **ht_cm:** Predicted height of the individual tree in centimeters.
    * **above_kg:** Above-ground tree biomass (including all wood, bark, branch, and foliage) in kilograms.
    * **leaf_kg:** Tree foliage biomass in kilograms.
    * Any additional (not required) columns in the input dataframe will remain in the output dataframe. 

* by_plot: 
    * **watershed:** As described above in external_data.
    * **year:** As described above in external_data.
    * **plot:** As described above in external_data.
    * **forest_type:** As described above in external_data.
    * **elev_m:** As described above in external_data.
    * **above_L_Mg_ha:** Above-ground live tree biomass (including all wood, bark, branch, and foliage) in megagrams per hectare.
    * **above_D_Mg_ha:** Above-ground dead tree biomass (including all wood, bark, branch) in megagrams per hectare.
    * **above_Mg_ha:** Above-ground (live + dead) tree biomass (including all wood, bark, branch, and foliage) in megagrams per hectare.
    * **leaf_Mg_ha:** Tree foliage biomass in megagrams per hectare.
  
* by_size: 
    * All columns described above in by_plot.
    * **sample_class:** Either tree (DBH >= 10 cm) or sapling (DBH < 10 cm).

### Demonstrations

```{r, include = FALSE}
library(HubbardBrookForestAnalytics)
```

**Results summarized by tree (using internal data):**
```{r}
# call the HBEFBiomass() function
internal_demo_1 <- HBEFBiomass(data_type = "internal",
                               results = "by_tree")

head(internal_demo_1)
```

*Notice in the output dataframe: The `bbd`, `canopy`, `elev_band`, `forest_type`, and `plot_area_ha` columns, which are additional (not required) columns in the internal dataframe, remain in the output dataframe. Any additional columns in the internal or external dataframe will remain in the output dataframe.*

<br>

**Results summarized by plot (using external data):**
```{r}
# investigate input external dataframe
external_demo_data
```

*Notice in the input dataframe: The plot without trees (W6, 1997, plot 2) is represented by species = "NONE". Watershed, year, plot, elev_m, and exp_factor are filled in as appropriate. All tree-level variables (i.e., status, vigor, dbh_cm) are NA.*

<br>

```{r}
# call the HBEFBiomass() function
internal_demo_2 <- HBEFBiomass(data_type = "external",
                               external_data = external_demo_data,
                               results = "by_plot")

internal_demo_2
```

*Notice in the output dataframe: The plot without trees (W6, 1997, plot 2) has 0 Mg/ha in all biomass columns.*

<br>

**Results summarized by plot as well as by size (using internal data):**
```{r}
# call the HBEFBiomass() function
internal_demo_3 <- HBEFBiomass(data_type = "internal",
                               results = "by_size")

head(internal_demo_3)
```

*Notice in the output dataframe: Since forest_type is included in the internal data, it is carried through into the summarized output data. The same idea applies to external data - if forest_type is included in the input dataframe, it will be carried through into the summarized output data.*

<br>

# Background information for tree biomass estimations

## Biomass workflow

**Step 1: Predict tree heights**

We fit new equations to predict height based on species, DBH, and elevation. These equations were fit using valley-wide height samples. 

**Step 2. Calculate above-ground live tree biomass**

We fit new equations to predict total above-ground tree biomass (including all wood, bark, branch, and foliage). We fit separate equations to predict tree foliage biomass on its own. 

**Step 3. Discount above-ground tree biomass for dead trees**

Standing dead trees lose mass in two ways: (1) they degrade with pieces falling and transferring to other biomass pools (e.g., stem tops break and become coarse woody debris) and (2) the remaining structures decay as measured by their density (i.e., mass/volume). We accounted for the decay and degrade of standing dead trees by discounting the live biomass estimates based on vigor. For trees/saplings with a vigor code of 4, the discount rate is 0.7283 (above-ground live tree biomass x 0.7283). For trees/saplings with a vigor code of 5, the discount rate is 0.5683 (above-ground live tree biomass x 0.5683). 


## Species code tables 

All hardwood and softwood species currently included/recognized in the `HBEFBiomass()` function are listed in the tables below. 

**Hardwoods** 

|common name|scientific name|4-letter code|
|:---|:---|:---|
|American beech|Fagus grandifolia|FAGR|
|Bigtooth aspen|Populus grandidentata|POGR|
|Black cherry|Prunus serotina|PRSE|
|Black ash|Fraxinus nigra|FRNI|
|Grey birch|Betula populifolia|BEPO|
|Juneberry/serviceberry|Amelanchier spp.|AMSP|
|American mountain-ash|Sorbus americana|SOAM|
|Mountain birch|Betula cordifolia|BECO|
|Mountain maple|Acer spicatum|ACSP|
|Paper birch|Betula papyrifera|BEPA|
|Pin cherry|Prunus pensylvanica|PRPE|
|Quaking aspen|Populus tremuloides|POTR|
|Red maple|Acer rubrum|ACRU|
|Sugar maple|Acer saccharum|ACSA|
|Striped maple|Acer pensylvanicum|ACPE|
|White ash|Fraxinus americana|FRAM|
|Yellow birch|Betula alleghaniensis|BEAL|
|Red elderberry|Sambucus racemosa|SAPU|
|Chokecherry|Prunus virginiana|PRVI|
|Alternateleaf dogwood|Cornus alternalternifolia|COAL|
|Basswood|Tilia americana|TIAM|
|Red oak|Quercus rubra|QURU|
|Eastern hophornbeam|Ostrya virginiana|OSVI|
|Willow species|Salix spp.|SASP|
|Cherry species|Prunus spp.|PRSP|
|Unknown|NA|UNKN|

<br>

**Softwoods**

|common name|scientific name|4-letter code|
|:---|:---|:---|
|Balsam fir|Abies balsamea|ABBA|
|Eastern hemlock|Tsuga canadensis|TSCA|
|Red spruce|Picea rubens|PIRU|
|White pine|Pinus strobus|PIST|
|Unknown|NA|UNKN|

*Note: Four-letter species codes are generally the first two letters of the genus followed by the first two letters of the species.*


## Vigor code table 

|vigor class|description|
|:---|:---|
|0|alive (Siccama code for live)|
|1|alive, healthy|
|2|alive, sick|
|3|alive, dying|
|4|standing dead tree|
|5|standing dead tree with at least 1/3 of top missing (stub)|
|6|dead and down (not recognized in R package)|

*Note: There is some historical variance in vigor codes. These are Cleavitt-Battles era codes.*

